# Reproducible Research: Peer Assessment 1

Course Project by Aleksei Neveror

```{r echo = TRUE}
library(ggplot2)
library(knitr)
opts_chunk$set(fig.path="figure/")
```

## Loading and preprocessing the data

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:

[Activity monitor data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

To perform the analysis we have to read sourse data
```{r echo = TRUE}
data <- read.csv("activity.csv")
```

And research the structure of the dataset
```{r echo = TRUE}
dim(data)
str(data)
```

To perform futher analysis we have to transform the dataset:

* change the type of the variable **data**
* create a new factor variable **daytype**, which should indicate is the day a weekday or weekend day. **weekend** and **weekday** are the possible values of this variable and also the factor levels

```{r echo = TRUE}
data$date <- as.Date(data$date)

data$daytype  <- factor(ifelse(as.POSIXlt(data$date)$wday %in% c(0,6),"weekend","weekday"))
```

After changes we should have this structure of the dataset

```{r echo = TRUE}
str(data)
```

## What is mean total number of steps taken per day?

To evaluate the total number of steps taken per day it's nessesary to aggregate data by date

```{r echo = TRUE}
totals <- aggregate(data$steps, by = list(data$date), FUN = sum)
names(totals) <- c("date","steps")
dim(totals)
head(totals)
```

The result dataset contains 61 rows and just 2 variables. As we can see the dataset **totals** contains NAs as values of the **steps** variable. Try to calculate the number of NAs values in this dataset:

```{r echo = TRUE}
sum(is.na(totals$steps))
mean(is.na(totals$steps))
```

There are just 8 NA values of the **steps** variable or 13,1% of values. For now we'll just take in attention that there are NAs values, but later we'll try to fill missing values in datasets.

Let's look through the total numbers of steps taken per day. For this purpose we have to calculate mean and median values of the total numbers of steps:

```{r echo = TRUE}
mean(totals$steps, na.rm =TRUE)
median(totals$steps, na.rm =TRUE)
```

These values means that the individual was wearing this activity monitor did circa 10766,19 steps per day, and with the 50%-probability he did less than 10765 steps per day. 

```{r echo = TRUE}
g <- ggplot(totals, aes(steps))
g <- g + geom_histogram(binwidth = 1000)
g <- g + labs(title = "Total number of steps taken each day")
print(g)
```

The histogram shows than the individual did most frequently between 10000-15000 per day.

## What is the average daily activity pattern?

To analyse the average daily activity pattern it's nessesary to aggregate data by interval calculating the mean values for each interval. By aggregating we ignore the NAs:

```{r echo = TRUE}
avgs <- aggregate(data$steps, by = list(data$interval), FUN = mean, na.rm = TRUE) 
names(avgs) <- c("interval", "steps")
avgs$interval <- as.numeric(paste(avgs$interval))
dim(avgs)
str(avgs)
```

The time series plot of the 5-minute interval and the average number of steps taken, averaged across all days, is shown below:
```{r echo = TRUE}
g <- ggplot(avgs, aes(interval, steps))
g <- g + geom_point(col = "blue", alpha = .5, size = 1)
g <- g + geom_path()
g <- g + scale_x_continuous(breaks = round(seq(min(avgs$interval), max(avgs$interval), by = 200),1))
g <- g + labs(title = "The average daily activity pattern")
g <- g + xlab("5-minutes intervals")
g <- g + ylab("Average number of steps")
print(g)
```

So we can see, that the individual took the most number of steps in the interval between 8:00 and 9:00. To determitane the 5-min interval in which the individual took the most number of steps it's nessesary to calculate the maximum of steps:

```{r echo = TRUE}
as.integer(avgs[avgs$steps == max(avgs$steps),"interval"])
```

So the 5-min interval in which the individual took the most number of steps is 8:35.


## Imputing missing values

Let's look at the source dataset. There are some obesrvations in which the **step** variable has NAs:

```{r echo = TRUE}
sum(is.na(data$steps))
mean(is.na(data$steps))
```

In fact there are 2304 NA values of the **step** variable (or 13,1%). It's not too a lot, but presenting of the NA values may introduce bias into some calculations or summaries of the data. 

To get rid of NA values let's replace them with some others. As value to replace let's use the mean value of steps taken by individual in the defined 5-minutes interval, calculated without NAs:

```{r echo = TRUE}
fdata <- data
fdata$steps <- as.numeric(fdata$steps)
fdata <- merge(fdata, avgs, by = "interval")
fdata <- transform(fdata, steps.x = ifelse(!is.na(steps.x),steps.x,steps.y))
fdata <- fdata[,1:4]
names(fdata)[2] <- "steps"
mean(is.na(fdata$steps))
```

As a result we have a new dataset with 0% of NA values in **steps** variable. 

For futher analisys we need to calculate new total values of numbers of steps per day:

```{r echo = TRUE}
ftotals <- aggregate(fdata$steps, by = list(fdata$date), FUN = sum)
names(ftotals) <- c("date","steps")
mean(is.na(ftotals$steps))
```

As we can see this dataset has 0% of NA values in **steps** variable too.

In order to estimate changes of the dataset let's plot two histograms and compare them:

```{r echo = TRUE}
## Merge datasets with and without NAs and mark the observations
tmp1 <- totals
tmp1$source <- "With NAs"
tmp2 <- ftotals
tmp2$source <- "Without NAs"
tmp <- rbind(tmp1, tmp2)
tmp$source <- factor(tmp$source)

### Plot a histogram
g <- ggplot(tmp, aes(steps))
g <- g + geom_histogram(binwidth = 1000, aes(color = source, fill = source))
g <- g + facet_grid(~source)
g <- g + labs(title = "Total number of steps taken each day")
print(g)
```

After inserting NA values the frequency of steps around 10000 per day has grown up.

Let's compare the mean and median values of old and new datasets:
```{r echo = TRUE}
# Summary for old dataset (with NAs)
summary(totals$steps)
# Summary for new dataset (without NAs)
summary(ftotals$steps)
```

The mean value didn't change after inserting NA values. But we can see some changes of the quantiles. So we can notice that the median value grew up. It shows us that individual takes 10770 or less steps per day with 50% probability. Also the 1st and the 3rd quantiles were changed: the 1st one grew up, and the 3rd one decreased. In compliance with new data the individual takes more steps with 25%-probability (9819 steps according the new dataset instead of 8841 steps according the old one) and less steps with 75%-probability (12810 steps per day according the new data instead of 13290 according the old one).

The analysis after imputting of empty values shows that the individual is less active than it was shown with the analisys of the dataset containing NAs, because with the higher probability the individual takes less steps per day.

## Are there differences in activity patterns between weekdays and weekends?

Using the new dataset (without NA values) let's build activity patterns for weekdays and weekends. For this purpose let's create a new dataset containing the average values of steps numbers per 5-minutes intervals based on the new dataset.This times we shall aggregate data by two parameters: interval and daytype:

```{r echo = TRUE}
favgs <- aggregate(fdata$steps, by = list(fdata$interval, fdata$daytype), FUN = mean, na.rm = TRUE)  
names(favgs) <- c("interval","daytype","steps")
favgs$interval <- as.integer(favgs$interval)
```

To compare activity patterns for weekdays and weekends let's plot the time series diagram:

```{r echo = TRUE}
g <- ggplot(favgs, aes(interval, steps))
g <- g + geom_point()
g <- g + geom_path()
g <- g + geom_smooth()
g <- g + facet_wrap( ~ daytype)
g <- g + labs(title = "Activity patterns for weekdays and weekends")
g <- g + xlab("5-minutes intervals")
g <- g + ylab("Average number of steps")
print(g)

```

The diagram shows that on the weekdays the individual takes more activity mornings (around 9:00). Later the number of taken steps significally decreases. Peak values are observed primarily in the morning and much lower in the evening. 

In comparison with activity pattern for weekdays the one for weekend is more flat. There isn't any significant picks, like in the pattern for weekdays. It shows that the individual is almost equally active throughout all the day at the weekends.
